---
title: "Covid19"
author: "Adrianna Queen"
date: "2023-03-05"
output:
  pdf_document: default
  html_document: default
---

Imports
```{r imports, include=FALSE}
library(tinytex)
library(tidyverse)
library(lubridate)
```



```{r cars}
filenames = 
  c("time_series_covid19_confirmed_US.csv",  "time_series_covid19_confirmed_global.csv", "time_series_covid19_deaths_US.csv",  "time_series_covid19_deaths_global.csv")

url_in = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"

urls = str_c(url_in,filenames)

global_cases = read_csv(urls[2])
global_deaths = read_csv(urls[4])
us_cases = read_csv(urls[1])
us_deaths = read_csv(urls[3])
```

After looking at global_cases and global_deaths, I would like to tidy those datasets and put each variable (date, cases, deaths) in their own column
Also, I don't need lat and lon for the analysis I am planning, so I will get rid of those and rename Region and State to be more r friendly

```{r pressure, echo=FALSE}
global_cases = global_cases %>%
  pivot_longer(cols=
                 -c('Province/State',
                    'Country/Region',Lat,Long),
              names_to="date",
              values_to="cases") %>%
  select(-c(Lat,Long))

#global_cases = global_cases %>% select(-c(Lat,Long))

global_deaths = global_deaths %>%
  pivot_longer(cols= -c('Province/State','Country/Region',Lat,Long),names_to='date',values_to='deaths') %>%
  select(-c(Lat,Long))
```

```{r Tidy and Transform}
global = global_cases %>%
  full_join(global_deaths) %>%
  rename(Country_Region = `Country/Region`,
         Province_State = `Province/State`) %>%
  mutate(date = mdy(date))

global = global %>% filter(cases > 0)

us_cases = us_cases %>%
  pivot_longer(cols= -(UID:Combined_Key),names_to="date",values_to="cases") %>%
                 select(Admin2:cases)%>%
                 mutate(date=mdy(date)) %>%
                 select(-c(Lat,Long_))

us_deaths = us_deaths %>%
  pivot_longer(cols=-(UID:Population),names_to='date',values_to='deaths') %>%
  select(Admin2:deaths) %>%
  mutate(date=mdy(date)) %>%
   

US = us_cases %>%
  full_join(us_deaths)

global = global %>%
  unite("Combined_Key",
        c(Province_State,Country_Region),
        sep=",",
        na.rm = TRUE,
        remove = FALSE)

uid_lookup_url = "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"

uid = read_csv(uid_lookup_url) %>%
  select(-c(Lat,Long_,Combined_Key,code3,iso2,iso3,Admin2))

global = global %>%
  left_join(uid,by=c('Province_State','Country_Region')) %>%
  select(-c(UID,FIPS)) %>%
  select(Province_State,Country_Region,date,cases,deaths,Population,Combined_Key)
```

Visualizing Data
```{r Visualizing Data}
us_by_state = US %>%
  group_by(Province_State,Country_Region,date) %>%
  summarize(cases=sum(cases),deaths=sum(deaths),
            Population=sum(Population)) %>%
  mutate(deaths_per_mill=deaths*1000000/Population) %>%
  select(Province_State,Country_Region,date,cases,deaths,deaths_per_mill,Population) %>%
  ungroup()

us_totals = us_by_state %>%
  group_by(Country_Region,date) %>%
  summarize(cases=sum(cases),deaths=sum(deaths),Population=sum(Population)) %>%
  mutate(deaths_per_mill=deaths*1000000/Population) %>%
  select(Country_Region,date,cases,deaths,deaths_per_mill,Population) %>%
  ungroup()

us_totals %>%
  filter(cases>0) %>%
  ggplot(aes(x=date,y=cases)) +
  geom_line(aes(color='cases')) +
  geom_point(aes(color='cases')) +
  geom_line(aes(y=deaths,color='deaths')) +
  geom_point(aes(y=deaths,color='deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom',
        axis.text.x=element_text(angle=90)) +
  labs(title='Covid19 in US',y=NULL)

state = "New York"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases>0) %>%
  ggplot(aes(x=date,y=cases)) +
  geom_line(aes(color='cases')) +
  geom_point(aes(color='cases')) +
  geom_line(aes(y=deaths,color='deaths')) +
  geom_point(aes(y=deaths,color='deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom',
        axis.text.x=element_text(angle=90)) +
  labs(title=str_c("Covid19 in ",state),y=NULL)
```


Analyzing Data
```{r Analyzing Data}
us_by_state = us_by_state %>%
  mutate(new_cases=cases-lag(cases),
         new_deaths=deaths-lag(deaths))
us_totals = us_totals %>%
  mutate(new_cases=cases-lag(cases),
         new_deaths=deaths-lag(deaths))

us_totals %>%
  filter(cases>0) %>%
  ggplot(aes(x=date,y=new_cases)) +
  geom_line(aes(color='new_cases')) +
  geom_point(aes(color='new_cases')) +
  geom_line(aes(y=new_deaths,color='new_deaths')) +
  geom_point(aes(y=new_deaths,color='new_deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom',
        axis.text.x=element_text(angle=90)) +
  labs(title='Covid19 in US',y=NULL)

state = "New York"
us_by_state %>%
  filter(Province_State == state) %>%
  filter(cases>0) %>%
  ggplot(aes(x=date,y=new_cases)) +
  geom_line(aes(color='new_cases')) +
  geom_point(aes(color='new_cases')) +
  geom_line(aes(y=new_deaths,color='new_deaths')) +
  geom_point(aes(y=new_deaths,color='new_deaths')) +
  scale_y_log10() +
  theme(legend.position = 'bottom',
        axis.text.x=element_text(angle=90)) +
  labs(title=str_c('Covid19 in',state),y=NULL)

us_state_totals = us_by_state %>%
  group_by(Province_State) %>%
  summarise(deaths=max(deaths),cases=max(cases),
            population=max(Population),
            case_per_thou=1000*cases/population,
            deaths_per_thou=1000*deaths/population) %>%
  filter(cases>0,population>0)

us_state_totals %>%
  slice_min(deaths_per_thou,n=10) %>%
  select(deaths_per_thou,case_per_thou,everything())
us_state_totals %>%
  slice_max(deaths_per_thou,n=10) %>%
  select(deaths_per_thou,case_per_thou,everything())
```

Modeling Data
```{r modeling}
mod = lm(deaths_per_thou ~ case_per_thou, data=us_state_totals)
x_grid = seq(140,450)
new_df = tibble(case_per_thou=x_grid)
us_tot_w_pred = us_state_totals %>% mutate(pred=predict(mod))

us_tot_w_pred %>% ggplot() +
  geom_point(aes(x=case_per_thou,y=deaths_per_thou),color='blue') +
  geom_point(aes(x=case_per_thou,y=pred),color='red')
```

## Bias Discussion
The Covid 19 Pandemic has lead to a significant amount of division between people. Some individuals believe at one extreme that Covid was a hoax and a government scam to control people, while on the other extreme people believe that the best thing to do is to social distance and stay home forever to help prevent as many covid19 cases as possible. This has led to many people feeling attached to their beliefs surrounding covid-19 and makes them more likely to want to prove their side is right through data.

It is important for the data scientist to remain as neutral as possible when doing analysis because even slight bias can negatively influence the results. One way they can do this is by getting inputs from people who fall on both sides of the political aisle, so that they can gain better context. Another way is to carefully word their hypotheses to be tested. If they word their hypotheses in a way that causes them to only look at certain parts of the data that would prove or disprove one thing, then they could miss important information that the data has to offer.